Vragenlijst thesis evaluatie

### 2015-06-10 doc/template/chap-10.tex
Tekst introduction bijgewerkt

A particular kind of threats are Advanced Persistent Threats (APT). \todo{misschien meer zeggen over APTs en threats} An APT is a targeted cyber attack that is designed to penetrate a network or a system in a stealthy way and that can stay undetected for a long period. This makes it so hard to protect a network or a system against an APT. Bruce Schneier describes an APT as something different and stronger than a conventional threat: ''\textit{A conventional hacker or criminal is not interested in any particular target. He wants a thousand credit card numbers for fraud, or to break into an account and turn it into a zombie, or whatever. Security against this sort of attacker is relative; as long as you're more secure than almost everyone else, the attackers will go after other people, not you. An APT is different; it's an attacker who - for whatever reason - wants to attack you. Against this sort of attacker, the absolute level of your security is what's important. It does not matter how secure you are compared to your peers; all that matters is whether you're secure enough to keep him out}'' - Bruce Schneier \cite{APTBruce}.\\

\subsubsection{Complication}
Since it is so difficult to protect a system or a network against APT's, researchers have been looking for effective ways to predict in advance which defence strategy might be the better one. 
Game theory is gaining increasing interest as an effective technique to model and study Cyber Security. Game theory analyses the security problem as a game where the players are an attacker and a defender of a system, and where both players have to make decisions. In particular, both players will aim for the strategy that results in a maximal benefit for them.  Researchers at RSA, van Dijk et al,  made a game theoretic model of targeted attacks. They study the specific scenario where a system or network is repeatedly taken over completely by an without being immediately detected by the defender of the system or network. In game theory, such a game is known as ''FlipIt'' \cite{FlipIt}. This is a two players game where the attacker and the defender are competing to get control over a shared resource. Both players do not know who is currently in control of the resource until they move. In FlipIt every move gives them immediate control over the resource. But what if the attacker moves and it takes a while before the attacker gets full control over the resource? FlipIt does not take into account that a move may not be instantaneous, but has a certain delay. Consider for example a network with different nodes ( laptops, datacenters) as a resource. The attacker drops a virus on one of the nodes and then wait till this virus infects the whole network. The attacker will only be in control of the resource once the whole network is infected. \\

@@

The game theoretical approach of the FlipIt does not take such delay into account. 
This an lead us to the following research questions:
\begin{itemize}
\item How can we incorporate the notion of delay in the game-theoretical analysis of the Flip-It game for a periodic strategy ?
\item Is there an optimal defense strategy against an attacker ? 
\end{itemize}

@@

\subsubsection{Overview of the thesis}
--opbouw van de thesis uitleggen-- \\

The organisation of this paper is the following.  In chapter \ref{Chapter1:Intro.Game.Theory}  a brief introduction to Gametheory is introduced to get familiar with the game theoretic concepts that will be further used in the paper. In the same chapter the FlipIt framework is summarized and the most important conclusions together with the the related work done on FlipIt and the difference with this paper.
In chapter \ref{chapter2:FlipIt with virus propagation} , we first introduce the adaptations made on FlipIt to model a FlipIt game with virus propagation. After that formulas are derived to model a FlipIt game with a virus propagation for a specific case where players play a periodic strategy with a random phase. This chapter ends with simulations where conclusions can be derived.
Next in Chapter \ref{chapter4: Virus propagation} we given an overview of the various ways in which a virus can propagate. We present a method to calculate the speed of the propagation of a virus in a network and how this network can be established to reduce the spreading of a virus.
Finally, in Chapter \ref{chapter5:conclusion} , we discuss the main results and complications and provide directions for further research.

>>>>

A particular kind of threats are Advanced Persistent Threats (APT). An APT is a multi-faceted, continuous and targeted cyber attack that is designed to penetrate a network or a system in a stealthy way and that can stay undetected for a long period. It is something different and stronger than a conventional threat. A conventional threat will not attack any particular target. An APT is persistent and will keep on trying to attack its victim. It operates silently and stealthily, to prevent detection. This makes it so hard to protect a network or a system against an APT.

%Bruce Schneier describes an APT as something different and stronger than a conventional threat: ''\textit{A conventional hacker or criminal is not interested in any particular target. He wants a thousand credit card numbers for fraud, or to break into an account and turn it into a zombie, or whatever. Security against this sort of attacker is relative; as long as you're more secure than almost everyone else, the attackers will go after other people, not you. An APT is different; it's an attacker who - for whatever reason - wants to attack you. Against this sort of attacker, the absolute level of your security is what's important. It does not matter how secure you are compared to your peers; all that matters is whether you're secure enough to keep him out}'' - Bruce Schneier \cite{APTBruce}.\\

\subsubsection{Complication}
Since it is so difficult to protect a system or a network against APT's, researchers have been looking for effective ways to predict in advance which defence strategy might be the better one. 
Game theory is gaining increasing interest as an effective technique to model and study cyber security problems. It is common to model cyber security games as a game with two players, an attacker and a defender. However, there are games that have more players \cite{fengstealthy}. In this paper we focus on the game with two players. The actions available to the attacker and the defender correspond to attacks on the system and defensive measures that protect the system. In particular, both players will execute the strategy that results in a maximal benefit for them.  Researchers at RSA, van Dijk et al,  made a game theoretic model of targeted attacks. They study the specific scenario where a system or network is repeatedly taken over completely by an without being immediately detected by the defender of the system or network. In game theory, such a game is known as ''FlipIt'' \cite{FlipIt}. This is a two players game where the attacker and the defender are competing to get control over a shared resource. Neither player knows who is currently in control of the resource until they move. In FlipIt every move gives them immediate control over the resource. But what if the attacker moves and it takes a while before the attacker gets full control over the resource? FlipIt does not take into account that a move may not be instantaneous, but has a certain delay. Consider for example a network with different nodes ( laptops, datacenters) as a resource. The attacker drops a virus on one of the nodes and then wait till this virus infects the whole network. The attacker will only be in control of the resource once the whole network is infected. \\

@@

This paper adapts the model presented in \cite{FlipIt} so as to take the delay for virus propagation into account.
This leads us to the following research questions:
\begin{itemize}
\item How can we incorporate the notion of delay in the game-theoretical analysis of the Flip-It game for a periodic strategy ?
\item Is there an optimal defense strategy against an attacker ? 
\end{itemize}

@@

\subsubsection{Overview of the thesis}

The organisation of this paper is the following.  In chapter \ref{Chapter1:Intro.Game.Theory}  a brief introduction to cyber security and gametheory is introduced to get familiar with the kind of threats that are in the scope of this work and the game theoretic concepts that will be further used in the paper. In the same chapter the FlipIt framework is summarized and the most important conclusions together with the the related work done on FlipIt and the difference with this paper.
In chapter \ref{chapter2:FlipIt with virus propagation} , we first introduce the adaptations made on FlipIt to model a FlipIt game with virus propagation. After that formulas are derived to model a FlipIt game with a virus propagation for a specific case where players play a periodic strategy with a random phase. \\
In Chapter \ref{chapter:Nash} the formules are used to determine if there is a nash equilibrium. \todo{aanpassen naar chapter met nash equilibria}
Next in Chapter \ref{chapter4: Virus propagation} we given an overview of the various ways in which a virus can propagate. We present a method to calculate the speed of the propagation of a virus in a network and how this network can be established to reduce the spreading of a virus.
Finally, in Chapter \ref{chapter5:conclusion} , we discuss the main results and complications and provide directions for further research.

### 2015-06-10 doc/template/chap-5.tex
Tekst introduction bijgewerkt

Before the digitalization of documents, information was kept on paper and the security of this information was ensured by administrative and physical means. For example, you needed a key to access documents stored in a room full of cabinets where the files were kept. In today's digital era more and more information is kept in a digital format, stored on a computer.  As digitalization progressed, the need for ensuring the security of digital information arose and automated tools where developed for protecting files stored on a computer.  The generic name to protect data stored on a computer controlled device such as computers and smartphones, as well as public and private computer networks, including the entire Internet is called computer security. \\

Security is a general term that encompasses several dimensions. More specifically, computer security has tree key objectives that are fundamental to computer security:
\begin{description}
\item Confidentiality: This assures that the confidential of private data is not disclosed or made available to users that do not have the authorization. 
\item Integrity: This assures that data can not be altered by an unauthorized individual.
\item Availability: This assures that data is always accessible and that the service is not denied to authorized individuals.
\end{description}

 the purpose of security is to give certainty that data will not be removed without authorization (Confidentiality) that the data is always accessible (Availability), and that data can not be read or altered by someone who does not have the authorization (Integrity).

 These are the 3 key attributes of security, also known as the CIA triad. The tree concepts are fundamental security objectives for the securing of data, information and computing services. \\

''Cyber security is the process of applying security tools to ensure confidentiality, integrity, and availability of data. Cybersecurity attempts to ensure the protection of assets, which includes data, desktops, servers, buildings, and most importantly, humans. The goal of cyber security is to protect data both in transit and at rest. Countermeasures can be put in place in order to increase the security of data. Some of these measures include, but are not limited to, access control, awareness training, audit and accountability, risk assessment, penetration testing, vulnerability management, and security assessment and authorization.'' \todo{tekst overgenomen van wikipedia} \\

In the context of cyber security, the terms ' threat and attack' are often used interchangeably, referring to more or less the same thing. They however have a slightly different meaning. A threat refers to anything that can breach the security and that can cause a possible harm. It is a possible danger that can exploit a vulnerability. An attack is an assault on computer security that comes from a threat. It is an intelligent act that tries deliberately to breach security through vulnerabilities.  \\

In order to keep a system secure, it is important to mitigate the possible threats. The possible threats can take many forms, the most commen being spam, malware, spoofing, phishing and  DDoS attacks. 
A security report of 2014 reveals that 61\% of the attacks on companies are caused by malware. This is a piece of malicious program that is designed to penetrate unprotected systems or computers, and getting there to sensitive information. Malware can be divided into different subclasses that are not mutually exclusive:

\begin{description}
\item Virus: This is a malicious piece of code that replicates and tries to spread to infect other systems or files. The "I love you" virus is an example of a virus that quickly spreads. This virus propagates via mail systems. If someone opens an email with "I love you" with the virus in the annex this virus spreads itself by sending a mail to everyone in your contact list. So the virus can multiply rapidly and eventually a business network might shut down by the heavy traffic. In this example, there is a need for human interaction to spread the virus. If no one opens the mail the virus can not infect other systems.
Unfortunately, there are also viruses that can spread without human interaction. These viruses are referred to as worms. A worm is also a computer program that replicates itself to spread to other computers. Via a computer network, copies of the worm can be forwarded without an intermediary. The worm will use vulnerabilities of the system to infect other computers.
The Stuxnetworm is a very famous worm. Initially this worm spread via infected USB sticks and from then it could spread through the Internet to other computers. The purpose of the Stuxnetworm was to harm the centrifuges in nuclear reactors. Many reactors have been infected. From the standpoint of the defender, it is very important to respond as quickly as possible so that the worm can not spread. 
\item Rootkit: A rootkit is designed to be stealthy and hide a set of programs installed on a system from methods of detection.  The programs have administrator or root privileges to that system, which makes it possible to access the functions and services of the operating system, change files, take control over the monitor processes and send and receive network traffic.  A rootkit can make changes to the system to keep itself concealed from detection.
\item Trojan: This is a malicious program that disguises itself as something normal, so that users won't be suspicious of installing it.  
\item Backdoors: Also know as trapdoor, is a whole in the system or program that allows access 
\end{description}

A company can take different measures to defend itself against malware. The threats caused by malware can be divided into three main categories: known threats (70\%), unknown threats (29\%) and advanced threats (1\%) as proposed by Kasperksy [\todo{toevoegen referentie naar kaspersky APT report}].

The known threats are the easiest to  defend oneself against. Standard malware protection tools like firewalls and virus scanners can keep these kind of malware out of the system. Installing protection against unkown is also relatively easy. Then the remaining 1\% are the advanced threats, also known as APT. APT's are ... 
In this paper we focus on targeted attacks. Targeted attacks are malware designed to attack exclusively an organisation.

An APT is a persistent targeted attack that tries to penetrate a network to harm it while staying unseen for a long period of time. The motive of an APT is mostly cyber espionage, stealing sensitive data, sabotage or some other kind of ideological attacks. Advanced Persistent Threat are called  'Advanced' for the fact that these attacks are well funded and that (usually) the attacker itself needs a great expertise to successfully penetrate a network. Not all APTs are technical advanced though. The attacker can also try to exploit existing vulnerabilities simply based in the hope that his target has not yet secured himself against these vulnerabilities. 'Persistent' refers to the fact that the attacker keeps on trying to break in to his victim. He will not give up. The attack can be over various years and different steps. The threats stands for the fact that an APT is attempting to break the protection of valuable data the leakage of which outside the secure environment may severely damage the organisation owning the data.

Some examples of the biggest most rare APTs are listed to get an understanding of what APTs are capable of and how long they can stay unseen. [\todo{site kaspersky apt}]

\subsubsection{Equation}
Equation is a complex cyberattack platform where the first known sample is from 2002, but it was only discovered 12 years later in 2014. This APT propagates through usb drives, cd or physical media. It will search for exploits and will self-replicate itself to spread the infection. The purpose of this virus is to steal data and cyberespionage.

\subsubsection{Regin}

\subsubsection{Flame}
Way of propagation through USB drives, LAN spreading. purpose cyber espionage.

\subsubsection{Black energy}
 purpose cyber espionage and DDoS, data wiping. prop usb lan

According to [\todo{APT report} ]  the damage of one successful target attack can exceed over 2.54 million dollar. a company needs a defense mechanism to defend itself against apt. Prevention more important than ..
een van de manieren om dat te doen is via game theorie.

Security belangrijk. verschillende malware. uitleggen malware. verwijzen naar kaspersky security raport. APT grote threats. voorbeelden van APT's. waarom gametheory gebruiken.

\section{Intro game theory}
Gametheory is a mathematical study to analyse interactions between independent and self-interested agents. To get an understanding of the most important concepts of game theory, a short introduction based on the work of 
\cite{leyton2008essentials} and \cite{Coursera} is given in section \ref{Cha1:briefintro}  . For a more detailed and full introduction to game theory, the reader is referred to 
\cite{leyton2008essentials}.  In section \ref{cha1:FlipItGame} an overview of the FlipIt game is given with the definitions and concepts that will be used throughout the paper. 
The last section \ref{ch1:extendedWork} will cover the extensions and additions already made on FlipIt.

%------------------------------------------------%
%            Intro Game Theory 					 %
%------------------------------------------------%
\section{A brief introduction in Game Theory}
\label{Cha1:briefintro}

@@

%John Nash speelde ook een grote rol in de geschiedenis van de speltheorie. Hij is een van de wiskundigen geweest die speltheorie geformaliseerd heeft. Het Nash evenwicht werd naar hem vernoemd. Een Nash evenwicht wordt gezien als een evenwicht tussen beide spelers zodat ze allebei de beste tactiek kiezen en niet meer veranderen als de andere van tactiek veranderen. John Nash breide de theorie over het Nash evenwicht in een paper nog uit met gemengde strategieën. In 1994 kreeg John Nash samen met twee andere wiskundigen gespecialiseerd op het vlak van speltheorie de Nobelprijs voor de economie op basis van hun prestaties in de niet-coöperatieve speltheorie. . Over John Nash is een prachtige film 
\subsection{Best response and Nash Equilibrium}
One of the solution concepts in Game Theory for non-cooperative games is a Nash Equilibrium that we will use in this paper. A Nash Equilibrium is a subset of outcomes that can be interesting to analyse a game. To define this concept we first introduce the concept of best response. The best response for a player is the action of a player that maximizes it's pay-off for any given action of the other player. We define $BR_{i}$ as the best response function for player \textit{i}. The best response for player \textit{1} is given by : $a_{1} = BR_{1}(a_{2})$. A 
For a Nash Equilibrium each player has a consist list of actions and each player's action maximizes his or her pay-off given the actions of the other players. Nobody has the incentive to change his or her action if an equilibrium profile is played. We have a Nash Equilibrium for the pair $(a_{1}^{*},a_{2}^{*})$ where $a_{1}^{*} = BR_{1}(a_{2}^{*})$ and $a_{2}^{*} = BR_{2}(a_{1}^{*})$\\

 \todo{optimal strategy uitleggen}

POSTCONDITION: Uitgelegd: Strategien, acties, strategien, spelers, rationeel, Nash, best response
%In game theory, the Nash equilibrium is a solution concept of a non-cooperative game involving two or more players, in which each player is assumed to know the equilibrium strategies of the other players, and no player has anything to gain by changing only their own strategy.[1] If each player has chosen a strategy and no player can benefit by changing strategies while the other players keep theirs unchanged, then the current set of strategy choices and the corresponding payoffs constitutes a Nash equilibrium. The reality of the Nash equilibrium of a game can be tested using experimental economics method.

@@

\subsubsection{List of terms}
In the following list a couple of terms that will be used throughout the paper.
\begin{description}
\item \textit{Players}: Players are referred as the ones who are the decision makers. It can be a person, a company or an animal.  (they will act rational )
\item \textit{Actions}: Every player has actions that he or he can do. 
\item \textit{Strategies}: A strategy is the combination of different actions. A pure strategy is only one action.
\item \textit{Utility function}: The utility function is the mapping of the level of happiness of an agent about the state of the world to natural numbers.

@@

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.5]{../../doc/template/Images/DefFlipit}
\caption{A representation of a FlipIt game where both players are playing periodically and at discrete time intervals. Every move or flip is indicated by a blue or orange circle. The attacker is represented in orange and plays with a period of $\delta_{A}=4$. The defender is represented in blue and plays with a period of $\delta_{D}=3$. The blue and orange rectangles represent the amount of time the respective player is in control of the resource.}
\label{fig:FLipItDefault}
\end{figure}

>>>>

Before the digitalization of documents, information was kept on paper and the security of this information was ensured by administrative and physical means. For example, you needed a key to access documents stored in a room full of cabinets where the files were kept. In today's digital era more and more information is kept in a digital format, stored on a computer.  As digitalization progressed, the need for ensuring the security of digital information arose and automated tools where developed for protecting files stored on a computer.  The generic name to protect data stored on a computer controlled device such as computers and smartphones, as well as public and private computer networks, including the entire Internet is called computer security. 
Security is a general term that encompasses several dimensions. More specifically, computer security has tree key objectives that are fundamental to computer security:

~~\\
 \textit{Confidentiality}: This assures that the confidential of private data is not disclosed or made available to users that do not have the authorization. \\
 \textit{Integrity:} This assures that data can not be altered by an unauthorized individual.\\
 \textit{Availability:} This assures that data is always accessible and that the service is not denied to authorized individuals.\\

 %The purpose of security is to give certainty that data will not be removed without authorization (Confidentiality) that the data is always accessible (Availability), and that data can not be read or altered by someone who does not have the authorization (Integrity). 
 These are the 3 key attributes of security, also known as the CIA triad. The tree concepts are fundamental security objectives for the securing of data, information and computing services. \\
Cyber security is the process of applying security tools to ensure confidentiality, integrity, and availability of data. It is an attempt to protect websites, servers, data, applications, operating systems and all assets that need protection in a computer system.  Some of these tools may include detection, identification or removal tools. A detection tool will determine if an infection has taken place and will trace the threat. An identification tool will try to identify the threat to be able to know how to remove it. The removal tool will once the threat has been identified, remove the threat from the system so that it cannot spread any further.

%Cybersecurity attempts to ensure the protection of assets, which includes data, desktops, servers, buildings, and most importantly, humans. The goal of cyber security is to protect data both in transit and at rest. Countermeasures can be put in place in order to increase the security of data. Some of these measures include, but are not limited to, access control, awareness training, audit and accountability, risk assessment, penetration testing, vulnerability management, and security assessment and authorization.'' \todo{tekst overgenomen van wikipedia} \\

\subsubsection{Threats to computer systems}

In order to keep a system secure and to be able to apply security tools, it is important to mitigate the possible threats. The possible threats can take many forms, the most common being spam, malware, spoofing, phishing and  DDoS attacks.  In the context of cyber security, the terms ' threat and attack' are often used interchangeably, referring to more or less the same thing. They however have a slightly different meaning. A threat refers to anything that can breach the security and that can cause a possible harm. It is a possible danger that can exploit a vulnerability. An attack is an assault on computer security that comes from a threat. It is an intelligent act that tries deliberately to breach security through vulnerabilities.  \\

The most noteworthy group and biggest group of threats to computer systems is malware. 
This is a piece of malicious software that is designed to penetrate unprotected systems or computers, with the intent to get sensitive information, destroy data, or compromise the confidentiality, integrity or availability of the data or applications of the victim. A security report of 2014 \cite{SurveyKaspersky} reveals that 61\% of the attacks on companies are caused by malware. For this reason this section will examine the categories of malware threats. Different types of malware exist such as virusses, worms, flooders, rootkits, bots, spyware, adware and many more. This broad range of different types can be classified into two main categories, the first one based on the propagation method that is used and the second one for the payload or the variety of actions that the malware performs. Propagation methods include virusses, worms and trojans. Payload includes flooders, rootkits, bots, spyware and adware. In this paper we focus on the category of propagation methods and not on the actions that the malware performs. We give a brief explanation of the three main propagation methods of malware. \todo{opsplitsing uit het boek van network security genomen}

\begin{description}
\item \textit{Virus}: This is a malicious piece of code that replicates itself and tries to spread in order  to infect other systems or files. A typical virus will attach itself to a program, or an executable content on a computer. The ''I love you'' virus is an example of a virus where the virus attached itself as an executable to a mail. To propagate it used the mail systems. If someone opens an email with the "I love you"  in the annex, the virus will spread itself by sending a mail to everyone in the victim's contact list. So the virus can multiply rapidly and eventually a business network might shut down by the heavy traffic. In this example, there is a need for human interaction to spread the virus. If no one opens the mail the virus can not spread itself and infect other systems.

\item \textit{Worm:} A worm is a virus that can spread without human interaction. A worm is a computer program that replicates itself in order to spread to other hosts on a network. Copies of the worm can be forwarded via a computer network without an intermediary. The worm will use vulnerabilities of the system to infect other computers.
The Stuxnetworm is a very prominent example of a worm. Initially this worm was spread via infected USB sticks and from there it could spread itself without any further human intervention through the Internet to other hosts on the network. The purpose of the Stuxnetworm was to harm the centrifuges in nuclear reactors and many reactors have been infected.

\item \textit{Trojan:} This is a malicious program that disguises itself as something normal and useful, so that users won't be suspicious of installing it, but it has a malicious function hidden inside that can avoid security measures and cause harm.  A notable trojan horse is Koobface, that targeted users of facebook, skype, yahoo, gmail and AOL mail. To spread itself the worm sended a mail or friend request with a message that directed the recipients to another third party website. This site tried to put the reciepient into downloading an update of Adobe Flash Player. Once downloaded and executed, Koobface could infect the host.

A company can take different measures to defend itself against malware. The threats caused by malware can be divided into three main categories: known threats (70\%), unknown threats (29\%) and advanced threats (1\%) as proposed by Kasperksy \cite{APTKaspersky}.

The known threats are the easiest to defend oneself against. Standard malware protection tools like firewalls and virus scanners can keep these kind of malware out of the system. Installing protection against unkown is also relatively easy, but this will need tools that go beyond the standard methods like dynamic whitelisting. The remaining 1\% are the advanced threats, also known as APT.

\subsubsection{Advanced threats to computer systems}
%\item Backdoors: Also know as trapdoor, is a whole in the system or program that allows access 
%\item Rootkit: A rootkit is designed to be stealthy and hide a set of programs installed on a system from methods of detection.  The programs have administrator or root privileges to that system, which makes it possible to access the functions and services of the operating system, change files, take control over the monitor processes and send and receive network traffic.  A rootkit can make changes to the system to keep itself concealed from detection.

 An APT is a persistent targeted attack that tries to penetrate a network to cause harm while staying unseen for a long period of time. The motive of an APT is mostly cyber espionage, stealing sensitive data, sabotage or some other kind of ideological attacks. Advanced Persistent Threat are called  'Advanced' for the fact that these attacks are well funded and that (usually) the attacker itself needs a great expertise to successfully penetrate a network. Not all APTs are technical advanced though. The attacker can also try to exploit existing vulnerabilities simply based in the hope that his target has not yet secured himself against these vulnerabilities. 'Persistent' refers to the fact that the attacker keeps on trying to attack his target. The attack can be over various years and different steps can be taken. The 'threats' stands for the fact that an intelligent act tries deliberately to breach security through vulnerabilities which may severely damage the target organisation. 
An APT can be a mix of different types of malware and will use different kinds of propagation methods. \\
%
%Some examples of the biggest most rare APTs are listed to get an understanding of what APTs are capable of and how long they can stay unseen. [\todo{site kaspersky apt}] 
%
%\subsubsection{Equation}
%Equation is a complex cyberattack platform where the first known sample is from 2002, but it was only discovered 12 years later in 2014. This APT propagates through usb drives, cd or physical media. It will search for exploits and will self-replicate itself to spread the infection. The purpose of this virus is to steal data and cyberespionage. 
%
%\subsubsection{Regin}
%
%\subsubsection{Flame}
%Way of propagation through USB drives, LAN spreading. purpose cyber espionage.
%
%\subsubsection{Black energy}
% purpose cyber espionage and DDoS, data wiping. prop usb lan

According to a security survey of Kaspersky \cite{SurveyKaspersky}  the damage of one successful target attack against a large company can exceed over 2.54 million dollar. A company needs a defence mechanism to defend itself against APTs. As previously said, detection and identification tools will not work against APTs. The removal tool will only work if the threat has been identified. To mitigate these kind of attacks there is a need for another security countermeasures. Here is where the researchers of RSA fall in. They have investigated through game theory how to model targeted attacks and what the best defend strategy is according to the information that is available.

\section{A brief introduction in Game Theory}
\label{Cha1:briefintro}
Gametheory is a mathematical study to analyse interactions between independent and self-interested agents. To get an understanding of the most important concepts of game theory, a short introduction based on the work of 
\cite{leyton2008essentials} and \cite{Coursera} is given in this section. For a more detailed and full introduction to game theory, the reader is referred to 
\cite{leyton2008essentials}.  \\
%In section \ref{cha1:FlipItGame} an overview of the FlipIt game is given with the definitions and concepts that will be used throughout the paper. 
%The last section \ref{ch1:extendedWork} will cover the extensions and additions already made on FlipIt. \\
%------------------------------------------------%
%            Intro Game Theory 					 %
%------------------------------------------------%

@@

%John Nash speelde ook een grote rol in de geschiedenis van de speltheorie. Hij is een van de wiskundigen geweest die speltheorie geformaliseerd heeft. Het Nash evenwicht werd naar hem vernoemd. Een Nash evenwicht wordt gezien als een evenwicht tussen beide spelers zodat ze allebei de beste tactiek kiezen en niet meer veranderen als de andere van tactiek veranderen. John Nash breide de theorie over het Nash evenwicht in een paper nog uit met gemengde strategieën. In 1994 kreeg John Nash samen met twee andere wiskundigen gespecialiseerd op het vlak van speltheorie de Nobelprijs voor de economie op basis van hun prestaties in de niet-coöperatieve speltheorie. . Over John Nash is een prachtige film 
\subsubsection{Best response and Nash Equilibrium}
One of the solution concepts in Game Theory for non-cooperative games is a Nash Equilibrium that will be used in this paper. A Nash Equilibrium is a subset of outcomes that can be interesting to analyse a game. To define this concept we first introduce the concept of best response. The best response for a player is the action of a player that maximizes it's pay-off for any given action of the other player. We define $Opt_{i}$ as the best response function for player \textit{i}. The best response for player \textit{1} is given by : $a_{1} = Opt_{1}(a_{2})$. A 
For a Nash Equilibrium each player has a consist list of actions and each player's action maximizes his or her pay-off given the actions of the other players. Nobody has the incentive to change his or her action if an equilibrium profile is played. We have a Nash Equilibrium for the pair $(a_{1}^{*},a_{2}^{*})$ where $a_{1}^{*} = Opt_{1}(a_{2}^{*})$ and $a_{2}^{*} = Opt_{2}(a_{1}^{*})$\\

\todo{optimal function same as best response ?}

%In game theory, the Nash equilibrium is a solution concept of a non-cooperative game involving two or more players, in which each player is assumed to know the equilibrium strategies of the other players, and no player has anything to gain by changing only their own strategy.[1] If each player has chosen a strategy and no player can benefit by changing strategies while the other players keep theirs unchanged, then the current set of strategy choices and the corresponding payoffs constitutes a Nash equilibrium. The reality of the Nash equilibrium of a game can be tested using experimental economics method.

@@

\subsubsection{List of terms}
In the following list a couple of terms that will be used throughout the paper.
\begin{description}
\item \textit{Players}: Players are referred as the ones who are the decision makers. It can be a person, a company or an animal.  (they will act rational )
\item \textit{Actions}: Every player has actions that he or she can do. 
\item \textit{Strategies}: A strategy is the combination of different actions. A pure strategy is only one action.
\item \textit{Utility}: The utility function or also known as the pay-of is the mapping of the level of happiness of an agent about the state of the world to natural numbers.

@@

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{../../doc/template/Images/FLipi.png}
\caption{A representation of a FlipIt game where both players are playing periodically. Every move or flip is indicated by a blue or orange circle. The attacker is represented in orange and plays with a period of $\delta_{A}=4$. The defender is represented in blue and plays with a period of $\delta_{D}=3$. The blue and orange rectangles represent the amount of time the respective player is in control of the resource.}
\label{fig:FLipItDefault}
\end{figure}
